---
title: "Ptarmigan-Abundance-Progress-Report"
author: "Rose An"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: TRUE
    toc_float: TRUE
link-citations: yes
fontsize: 12pt
bibliography: #Bibliography_Reproducible_Science_9.bib
csl: #AmJBot.csl
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
# Clear working environment
rm( list = ls() )

# Get working directory (set to right one if it's wrong)
getwd()
```

# Introduction and Objective

Willow and Rock Ptarmigan are primary prey sources for the Gyrfalcon, a declining Arctic raptor species. For this research project, I will aim to estimate annual ptarmigan abundance to support PhD student Michael Henderson's research linking adult survival of Gyrfalcons to ptarmigan numbers in Alaska along the Seward Peninsula.

```{r packages, include = FALSE, message = FALSE, warning = FALSE}
###~~~
# Load R packages
###~~~
#Create a vector w/ the required R packages
# --> If you have a new dependency, don't forget to add it in this vector
pkg <- c("knitr", "rmarkdown", "bookdown", "formattable", "kableExtra", "dplyr", "magrittr", "prettydoc", "htmltools", "knitcitations", "bibtex", "devtools", "tidyverse", "ggplot2", "raster", "unmarked")

##~~~
#2. Check if pkg are already installed on your computer
##~~~
print("Check if packages are installed")
#This line below outputs a list of packages that are not installed. Which ones of the packages above are installed in computer, print packages that are not installed. Should print 0.
new.pkg <- pkg[!(pkg %in% installed.packages())]

##~~~
#3. Install missing packages
##~~~
# Use an if/else statement to check whether packages have to be installed
# WARNING: If your target R package is not deposited on CRAN then you need to adjust code/function
if(length(new.pkg) > 0){
  print(paste("Install missing package(s):", new.pkg, sep=' '))
  install.packages(new.pkg, dependencies = TRUE)
}else{
  print("All packages are already installed!")
}

##~~~
#4. Load all required packages
##~~~
print("Load packages and return status")
#Here we use the sapply() function to require all the packages
# To know more about the function type ?sapply() in R console
# Just an easier way to load all packages
sapply(pkg, require, character.only = TRUE)
# debug = right-pointing arrow next to code will load in console so you can go line by line and debug. pressing this also seemed to resolve the CRAN needs a mirror error
# Generate BibTex citation file for all loaded R packages
# used to produce report Notice the syntax used here to
# call the function
knitr::write_bib(.packages(), file = "packages.bib")
# The .packages() function returns invisibly the names of all packages loaded in the current R session (to see the output, use .packages(all.available = TRUE)). This ensures that all packages used in your code will have their citation entries written to the .bib file. This packages is then used in the Appendix to cite the code used
```

```{r chunk setup, include = FALSE}
### Chunk options: see http://yihui.name/knitr/options/ ###
### Text output
opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE, include = TRUE)
## Code formatting
opts_chunk$set(tidy = TRUE, tidy.opts = list(blank = FALSE, width.cutoff = 60),
    highlight = TRUE)
## Code caching
opts_chunk$set(cache = 2, cache.path = "cache/")
## Plot output The first dev is the master for the output
## document
opts_chunk$set(fig.path = "Figures_Ptarm/", dev = c("png", "pdf"),
    dpi = 300)
## Figure positioning
opts_chunk$set(fig.pos = "H")
opts_chunk$set( dplyr.width = Inf, dplyr.print_min = 100 )
```

# Data

Since 2016, the Peregrine Fund has conducted annual distance sampling surveys along 10 mi road transects in the Seward Peninsula of Alaska. Each transect was surveyed twice yearly. Surveys were completed by two observers scanning for ptarmigan in a vehicle. When ptarmigan were spotted, observers would tally birds and record distance to the bird with rangefinders. Data from 2016-2026 was cleaned and combined into one csv by Kara Beer. The first rows of the dataset and its dimensions are shown here:

```{r pulldata, echo = FALSE}
# set path to where you can access your data #
# Note that the path will be different for your.#
datapath <- datadir <- paste( getwd(), "/Data/", sep = "" )
# load ptarmigan raw detection data
detdf <- read.csv( file = paste( datadir, "2026-07-23_all_prey_data.csv", sep = ""),
                   header = TRUE )
#view
head( detdf ); dim( detdf )
```
## Cleaning

```{r dataclean, include = FALSE}

#database has observations that are not ptarmigan. view all the species in the species column
unique(detdf$species)

#new dataframe with only relevant columns and only for ptarmigan obs.
cleandetdf <- detdf %>% 
  dplyr::select(year, round, observer1, observer2, species, cluster_size, distance) %>% #select relevant columns (cluster_size = ptarmigan count),
  dplyr::filter(., species == "WIPT" | species == "ROPT" | species == "PTAR") %>% #filter out rows that are not ptarmigan counts
  dplyr::mutate(., obsv_pair = paste(observer1, observer2, year, sep = ", ")) #create a new column with unique observer pairs and years

#view
head(cleandetdf); dim(cleandetdf)
#5263 observations
```
After filtering out observations of non-ptarmigan species, I am left with `r dim(cleandetdf)` observations.

## Visualization

Key to distance sampling models is fitting distance sampling curves to detection of ptarmigan by distance. First, I will create histograms of distance to ptarmigan by observer pair/year and survey to determine the type of curve that will be used in the model. 

```{r datavis, echo = FALSE}

#create an object that stores all unique observer pairs by year
pairs <- c(unique(cleandetdf$obsv_pair))
#view
pairs

#for loop to create histograms of distance by observer pair/year and survey (1st or 2nd)
for( p in 1:length(pairs) ){
  # create an object with the ggplot so that you can display it 
  # in a loop 
  a <- ggplot( subset(cleandetdf, obsv_pair %in% pairs[p]), aes(x = distance, fill = species) ) + #choose your data
    theme_bw( base_size = 15 ) + #choose a preset theme
    labs( title = pairs[p] ) + #label x axis using our observer pairs
    geom_histogram( stat = "count") + #plot histogram
    facet_wrap(~round) #plot by survey 1 and 2
    #scale_x_discrete(breaks = scales::pretty_breaks(5))
  # display your plot object
  print( a )
}

#distbins <- seq( min( cleandetdf[,"distance"]), max( cleandetdf[,"distance"]),
                  #length.out = 10 )
#unique(detdf$distance)
#str(detdf$distance)
#oh distances are characters bc of <100, etc.
```

```{r stopsave}
#save rdata and load in at next work session
save.image( "PtarmAbun.RData" )
load( "PtarmAbun.RData" )
```

# Analysis

I will use the unmarked package and gdistsamp or distsamp functions.